{\small} CUBEP3M: High Performance P$^{3}$M N-body code

### 2012-05-08 Halos.tex
More improvements
Updated references, Intro, scaling, halos with Ilian's input.

\section{Sperical Overdensity Halo Finder}

Spherical overdensities, search algorithm, 
provided halo information, Comparison to PS and ST.

>>>>

\section{Runtime Halo Finder}
\label{sec:halo}

%Spherical overdensities, search algorithm, 
%provided halo information, halo bias, comparison to PS and ST.
%{\bf (Ilian, You can lead the way here...)}

We have implemented a halo finding procedure, which we have developed 
based on the spherical overdensity (SO) approach \citep{1994MNRAS.271..676L}.
In the interest of speed and efficiency the halo catalogues are constructed 
on-the-fly at a pre-determined list of redshifts. The halo finding is 
massively-parallel and threaded based on the CubeP$^3$M data structures 
discussed earlier. The code first builds the 
fine-mesh density for each sub-domain using CIC or NGP interpolation. It then 
proceeds to search for and record all local density maxima above certain
threshold (typically set to 100 above mean density) within the local 
sub-domain's physical volume (excluding the sub-domain buffer zones). It then 
uses parabolic interpolation on the density field to determine more precisely
the location of the maximum within the densest cell, and records the peak 
position and value. The halo center determined this way agrees closely with 
the center-of-mass of the halo particles discussed below.

Once the list of peak positions is generated, they are sorted from the highest 
to the lowest density value. Then each of the halo candidates is inspected 
independently, starting with the highest peak. The grid mass is accumulated 
in spherical shells of fine grid cells surrounding the maximum, until the 
mean density within the halo drops below a pre-defined overdensity cutoff 
(usually set to 178 in units of the mean, in accordance to the top-hat 
collapse model). As we accumulate mass we remove it from the mesh, so that no 
mass element is double-counted. This method is thus inappropriate for finding 
sub-halos as within this framework those are naturally incorporated in their 
host halos. Because the halos are found on a grid of finite-size cells, it is 
possible, especially for the low-mass halos, to overshoot the target overdensity.
When this occurs we use an analytical halo density profile to correct the 
halo mass and radius to the values corresponding to the target overdensity. 
This analytical density profile is given by Truncated Isothermal Sphere (TIS) 
profile \citep{1999MNRAS.307..203S,2001MNRAS.325..468I} for overdensities below 
$\sim130$, and $1/r^2$ for lower overdensities. The TIS density profile has a
similar outer slope (the relevant one here) to the Navarro, Frenk and White 
\citep[NFW][]{1997ApJ...490..493N} profile, but extends to lower overdensities
and matches well the virialization shock position given by the Bertschinger 
self-similar collapse solution \citep{1985ApJS...58...39B}.

Once the correct halo mass, radius and position are determined, we find all 
particles which are within the halo radius. Their positions and velocities are
used to calculate the halo center-of-mass, bulk velocity, internal velocity 
dispersion and the three angular momentum components, all of which are then 
included in the final halo catalogues. We also calculate the total mass of
all particles within the halo radius, also listed in the halo data. This mass
is very close, but  typically slightly lower, than the halo mass calculated 
based on the gridded density field. The particle center-of-mass corresponds 
very well to the halo centre found based on the gridded mass distribution.

%Compared to Tinker
%44% low for 50 particles
%22% low for 400 particles
%12% low for 1000 particles

\begin{figure}%[ht]
%  \vskip -0.5cm 
  \begin{center}
    \includegraphics[width=3.2in]{graphs/mf_z0_Tinker.eps}
  \end{center}
  \caption{Simulated halo multiplicity function, 
    $\frac{M^2}{\bar{\rho}}\frac{dn}{dM}$ based on a
    CubeP$^3$M simulation with $3.2\,h^{-1}Gpc$ box and $4000^3$ 
    particles (red, solid). For reference we also show a widely-used 
    precise fit by \citet{2008ApJ...688..709T} (blue, dashed). 
    \label{mf}}
\end{figure}

A sample halo mass function produced based on our inlined SO halo finder 
at redshift $z=0$ from a simulation with $3.2\,h^{-1}Gpc$ box and $4000^3$
particles is shown in Figure~\ref{mf}. We compare our result to the 
precise fit presented recently by \citet{2008ApJ...688..709T}. Unlike most
other widely-used fits like the one by Sheth and Tormen 
\citep{2002MNRAS.329...61S}, which are based on friends-of-friends (FOF)
halo finders, the \citet{2008ApJ...688..709T} fit is based on a Spherical 
Overdensity-determined halo masses, which are systematically different 
from the FOF masses \citep[e.g.][]{2007MNRAS.374....2R,2008ApJ...688..709T}, 
making this fit a better base for comparison here. Results show excellent
agreement, within $\sim10\%$ for all halos with masses corresponding to
1000 particles or more. Lower-mass halos are somewhat under-counted compared
to the \citet{2008ApJ...688..709T} fit, by $\sim20\%$ for 400 particles and 
by $\sim40\%$ for 50 particles. This is due to the grid-based nature of our
SO halo finder, which misses some of the low-mass halos. Using more sophisticated
halo finders (available through post-processing due to their heavier memory
footprint) recovers the expected mass function.

{\bf II: do we need to say anything more about having the option to use other
halo finders like AHF and FOF?}

%{\bf Show some examples and comparisions to analytical fits and other halo finders.}

### 2012-05-08 Initial.tex
Included Ilian's and Vincent's first contribution (Halofinder and f_nl respectively).
Updated references, Intro, scaling, halos with Ilian's input.

As mention in section \ref{sec:structure}, the code start off by reading a set of initial conditions.
These correspond to a  $6 \times N$ phase-space array, where $N$ is the number of particles in the
local node. Although most applications so far were based on initial conditions that follow Gaussian statistics,
we have developed a non-Gaussian initial condition generator that we briefly describe in this section. 
{\bf Vincent, here you go!}

>>>>

As mention in section \ref{sec:structure}, the code start off by reading a set of initial conditions.
These correspond to a  $6 \times N$ phase-space array, where $N$ is the number of particles in the
local node. Although most applications so far were based on initial conditions that follow Gaussian statistics,
we have developed a non-Gaussian initial condition generator that we briefly describe in this section.

The original code that provides Gaussian initial conditions for {\small CUBEP3M} 
is extended to include non-Gaussian features of the ``local'' form,
$\Phi({\bf x})=\phi({\bf x})+f_{\rm NL} \phi({\bf x})^2 + g_{\rm NL} 
\phi({\bf x})^3$, where $\phi({\bf x})$ is the Gaussian contribution
to the Bardeen potential $\Phi({\bf x})$ (see \cite{F_NL} for a review). 
We adopted the CMB convention,
in which $\Phi$ is calculated immediately after the matter-
radiation equality (and not at redshift $z=0$ as in the large scale
structure convention). For consistency, $\phi({\bf x})$ is normalized
to the amplitude of scalar perturbations inferred by CMB measurements
($A_s\approx 2.2 \times 10^{-9}$). The local transformation is performed 
before the inclusion of the matter transfer function, and the initial 
particle positions and velocities are finally computed from $\Phi({\bf x})$ 
according to the Zel'dovich approximation, as in the original Gaussian initial condition generator.

This code was tested by comparing simulations and theoretical predictions
for the effect of local primordial non-Gaussianity on the halo mass 
function and matter power spectrum (Desjacques, Seljak \& Iliev 2009). 
It has also been used to quantify the impact of local non-Gaussian initial
conditions on the halo power spectrum \citep{2009MNRAS.396...85D,
2010PhRvD..81b3006D} and bispectrum \citep{2010MNRAS.406.1014S},
 as well as the matter bispectrum \citep{2011arXiv1111.6966S}.

### 2012-05-08 Introduction.tex
Updated references, Intro, scaling, halos with Ilian's input.

Many physical and astrophysical systems are subject to non-linear dynamics
and rely on N-body simulations to describe the evolution of bodies. 
One of the main field of application is the modelling of large scale structures, 
which are driven by the sole force of gravity. Recent observations of the cosmic microwave background
\citep{WMAP5-7} of galaxy clustering \citep{2000AJ....120.1579Y, 2003astro.ph..6581C, 2009arXiv0902.4680S, 2010MNRAS.401.1429D} of weak gravitational lensing \citep{CFHTLS, SDSS}
and of supernovae redshift-distance relations all point towards a standard model of cosmology, in which dark energy
and collisionless dark matter occupy more than 95 per cent of the total energy density of the universe.
In such a paradigm, pure N-body code are perfectly suited to describe the dynamics, as long as we
understands where and how the baryonic fluid feeds back on the dark matter structure.
The next generation of measurements aim at constraining the cosmological parameters at the per cent level, 
and the theoretical understanding of the non-linear dynamics that govern structure formation heavily relies 
on numerical simulations.

For instance, a measurement of the baryonic acoustic oscillation (BAO) dilation scale can provide
tight constraints on the dark energy equation of scale \citep{Seo2003,2005}.
The most optimal estimates of the uncertainty requires the knowledge of the 
matter power spectrum covariance matrix, which is only accurate when measured from a large sample 
of N-body simulations \citep{2005MNRAS.360L..82R, 2009ApJ...700..479T, 2011ApJ...726....7T}.
For the same reasons, the most accurate estimates of weak gravitational lensing signal is obtained
by propagating photons in past light cones that are extracted from simulated density fields \citep{2003ApJ...592..699V, 2009ApJ...701..945S, 2009A&A...499...31H}.
{\bf (Ilian, could you say something about reionization here? Just general references as to why one needs N-body codes to perform 
accurate predictions/forecasts, etc.)}

@@

With the advent of large computing facilities, parallel coding has now become common practice, 
and N-body codes have evolved both in performance and complexity. 
Many codes have opted for a tree algorithm, including Gadjet \citep{Gadjet, Gadjet2}, PMT \citep{1995ApJS...98..355X}, GOTPM \citep{2004NewA....9..111D}, and Hydra \citep{Hydra}, in which the local resolution increases with the density of the matter field. 
These have the advantage to balance the load across the computing units, which enable the calculation of high density regions. 
The drawback is a significant loss in speed, which can be only partly recovered by turning off the tree algorithm. 
The same reasoning applies to the mesh-refine code of  \cite{1991ApJ...368L..23C} and the PM code by \cite{FerrelBertschinger1995}.
 these are not designed to perform large scale PM calculations.

@@

This paper aims at presenting and quantifying these new features that make {\small CUBEP3M} one of the most competitive public N-body code.
It has already been involved in a number of scientific applications over the last few years,
spanning the field of weak lensing \citep{Vafaei10, 2008MNRAS.388.1819L,  2009arXiv0905.0501D, 2010PhRvD..81l3015L, 2010arXiv1012.0444Y, Harnois_Vafaei},  BAO
 \citep{2010arXiv1008.3506Z,  2011arXiv1106.5548N, 2011arXiv1109.5746H, HarnoisYuReco}, re-ionization \citep{Ilian...}, {\bf (anymore cubep3m papers to cite?)},  
and it seems relevant for the community to have access to a paper that describes the methodology, the accuracy and the performance of this public code 
The paper is structured as follow: XXXX

>>>>

Many physical and astrophysical systems are subject to non-linear dynamics
and rely on N-body simulations to describe the evolution of bodies. 
One of the main field of application is the modelling of large scale structures, 
which are driven by the sole force of gravity. Recent observations of the 
cosmic microwave background \citep{WMAP5-7}, of galaxy clustering 
\citep{2000AJ....120.1579Y, 2003astro.ph..6581C, 2009arXiv0902.4680S, 
2010MNRAS.401.1429D} of weak gravitational lensing \citep{CFHTLS, SDSS}
and of supernovae redshift-distance relations all point towards a standard 
model of cosmology, in which dark energy and collisionless dark matter occupy 
more than 95 per cent of the total energy density of the universe. In such a 
paradigm, pure N-body code are perfectly suited to describe the dynamics, as 
long as baryonic physics is not very important, or at least we understand how 
the baryonic fluid feeds back on the dark matter structure. The next generation 
of measurements aim at constraining the cosmological parameters at the per cent 
level, and the theoretical understanding of the non-linear dynamics that govern 
structure formation heavily relies on numerical simulations.

For instance, a measurement of the baryonic acoustic oscillation (BAO) dilation 
scale can provide tight constraints on the dark energy equation of scale 
\citep{Seo2003,2005}. The most optimal estimates of the uncertainty requires the 
knowledge of the matter power spectrum covariance matrix, which is only accurate 
when measured from a large sample of N-body simulations \citep{2005MNRAS.360L..82R, 
2009ApJ...700..479T, 2011ApJ...726....7T}. For the same reasons, the most accurate 
estimates of weak gravitational lensing signal is obtained by propagating photons 
in past light cones that are extracted from simulated density fields 
\citep{2003ApJ...592..699V, 2009ApJ...701..945S, 2009A&A...499...31H}.
Another area where large-scale N-body simulations have in recent years been 
instrumental are in simulations of early cosmic structures and reionization 
\citep[e.g.][]{2006MNRAS.369.1625I,2007ApJ...654...12Z,2007ApJ...671....1T,
2011arXiv1107.4772I}. The reionization process is primarily driven by low-mass 
galaxies, which for both observational and theoretical reasons, need to be resolved 
in fairly large volumes, which demands simulations with a very large dynamic range.   
%{\bf (Ilian, could you say something about reionization here? Just general references as to why one needs N-body codes to perform 
%accurate predictions/forecasts, etc.)}

@@

With the advent of large computing facilities, parallel coding has now become 
common practice, and N-body codes have evolved both in performance and complexity. 
Many codes have opted for a tree algorithm, including Gadget \citep{Gadjet, Gadjet2}, PMT \citep{1995ApJS...98..355X}, GOTPM \citep{2004NewA....9..111D}, and Hydra \citep{Hydra}, in which the local resolution increases with the density of the matter field. 
These have the advantage to balance the load across the computing units, which enable the calculation of high density regions. 
The drawback is a significant loss in speed, which can be only partly recovered by turning off the tree algorithm. 
The same reasoning applies to the mesh-refine code of  \cite{1991ApJ...368L..23C} and the PM code by \cite{FerrelBertschinger1995}.
 these are not designed to perform large scale PM calculations.

@@

This paper aims at presenting and quantifying these new features that make {\small CUBEP3M} one of the most competitive public N-body code.
It has already been involved in a number of scientific applications over the last few years,
spanning the field of weak lensing \citep{Vafaei10, 2008MNRAS.388.1819L,  2009arXiv0905.0501D, 2010PhRvD..81l3015L, 2010arXiv1012.0444Y, Harnois_Vafaei},  BAO
 \citep{2010arXiv1008.3506Z,  2011arXiv1106.5548N, 2011arXiv1109.5746H, HarnoisYuReco}, 
formation of early cosmic structures \citep{2008arXiv0806.2887I,2010arXiv1005.2502I},
observations of dark stars \citep{2010MNRAS.407L..74Z,2012MNRAS.tmp.2794I} and
reionization \citep{2011arXiv1107.4772I,Fernandez:2011ab,2011MNRAS.413.1353F,2012MNRAS.422..926M,Datta:2011hv,2012arXiv1203.0517F}, {\bf (anymore cubep3m papers to cite?)},  
and it seems relevant for the community to have access to a paper that describes the methodology, the accuracy and the performance of this public code 
The paper is structured as follow: XXXX

### 2012-05-08 ScalingPerformance.tex
Updated references, Intro, scaling, halos with Ilian's input.

Real time versus size, bottle neck, description of the largest runs, etc.
{\bf (Ilian, I will need your input here as well. Hugh, do you have anything you would like to share 
concerning the largest runs you did on Blue Gene? )}

>>>>

Real time versus size, bottle neck, description of the largest runs, etc.
{\bf (Hugh, do you have anything you would like to share 
concerning the largest runs you did on Blue Gene? )}

\begin{figure*}%[ht]
%  \vskip -0.5cm 
  \begin{center}
    \includegraphics[width=3.2in]{graphs/scaling_cubep3m_curie.eps}
    \includegraphics[width=3.2in]{graphs/scaling_cubep3m_new.eps}
%  \vskip -1.2cm 
%  \vskip -0.5cm 
  \caption{Scaling of {\small CUBEP$^3$M} on Curie fat nodes (left) and 
    on Ranger TACC facility for very large number of cores (right). Plotted is the code speedup 
    ($N_{\rm particles}^3/t_{\rm wallclock}$) against core count, normalized by the smallest run 
    in each case. Dashed line indicates the ideal weak 
    scaling. The data is listed in Table \ref{summary_scaling_table}.
    \label{scaling}
% \vskip -0.9cm 
}
\end{center}
\end{figure*}

\begin{table*}%[ht]
  \vskip -0.5cm 
  \begin{center}
\caption{Scaling of {\small CUBEP$^3$M} on Curie. Speedup is 
scaled to the smallest run.}
\label{summary_scaling_table}
\begin{tabular}{@{}|llllll|}
\hline
number of cores & speedup & ideal speedup & absolute timing (min) & 
$N_{\rm particles}$& box size (Mpc$/h$)
\\[2mm]\hline
%\hline
32  &  1.00 & - &3.2 & $256^3$ & 256\\
256  & 7.21 & 8 &3.55 & $512^3$  & 512\\
864  & 25.63 & 27 &4.8 & $864^3$  & 864\\
2048  & 61.87 & 64 &26.48 & $2048^3$ & 2048 \\
\hline
\end{tabular}
\caption{Scaling of the CubeP$^3$M code for large number of cores 
(on Ranger). Speedup is scaled to the smallest run.}
\label{summary_scaling_table2}
\begin{tabular}{@{}|llllll|}
\hline
number of cores & speedup & ideal speedup & absolute timing (min) & 
$N_{\rm particles}$& box size (Mpc$/h$)
\\[2mm]\hline
%\hline
864    & 1.00  & -    &258   & $1728^3$  & 6.3\\
4096   & 4.53  & 4.74 &320   & $3072^3$  & 11.4\\
10976  & 9.78  & 12.7 &845   & $5488^3$  & 20\\
21952  & 14.73 & 25.4 &561   & $5488^3$  & 20 \\
\hline
\end{tabular}
\end{center}
  \vskip -0.7cm 
\end{table*}

%the Ranger system at the Texas Supercomputing
%Center (in top 20 in the world) which is a SunBlade x6420 
%with AMD x86\_64 Opteron Quad Core 2300 MHz (9.2 GFlops)
% ``Barcelona'' processors and Infiniband networking. It has 
%a total of 62976 computing cores and 125952 GB of total 
%memory. Its nodes consist of 4 Quad Core processors and 32 GB 
%of shared RAM. For efficiency reasons (local memory access) we 
%typically use smaller MPI 'nodes' consisting of one Quad Core 
%processor and 8 GB of RAM.

The parallel algorithm of {\small CUBEP$^3$M} is designed for ``weak'' 
scaling, i.e. if the number of cores and the problem size 
increase in proportion to each other, then for ideal scaling the 
wall-clock time should remain the same, in contrast to ``strong'' 
scaling, whereby the same problem solved on more cores should take 
proportionately less wall-clock time. This weak scaling requirement 
is dictated by the problems we are typically aiming towards (very 
large and computationally-intensive) and our goals, which are to 
address such large problems in the most efficient way, rather than 
for the least wall-clock time. Furthermore, there is no explicit 
load balancing, thus the code is most efficient if the sub-domains
contain roughly equal number of particles, which is true for most
cosmological-size volumes, but not for e.g. simulations of a single
highly-resolved galaxy.

%Since the current Curie is using Intel Nehalem architecture, while 
%the thin Curie nodes will use the newer Westermere Intel architecture, 
%we also show the scaling of our code on the Westermere-based Lonestar 
%computer at the Texas Advanced Computing Centre

The scaling was tested with a dedicated series 
of simulations with increasing size and number of cores on the `fat'' 
(i.e. large-memory) nodes of the supercomputers Curie at TGCC in France. 
Our results are shown in Fig. \ref{scaling} and in 
Table \ref{summary_scaling_table}. For appropriate direct comparison,
all simulations were performed using the same particle mass 
($M_{\rm particle}=1.07\times10^{11}M_\odot$) and force resolution 
(softening length 50 $h^{-1}$kpc). The box sizes used range from 256 $h^{-1}$Mpc
to 2048 $h^{-1}$Mpc, and the number of particles from $256^3$ to $2048^3$.
Simulations were run on 32 up to 2048 computing cores starting from 
redshift $z=100$, and eveolving until $z=0$. Our results show excellent scaling, within 
$\sim3\%$ of the ideal one, for up to 2048 cores.

We have also ran {\small CUBEP$^3$M} on a much larger number of cores, 
from 8000 to up to 21,976, with $5488^3$-$6000^3$ (165 to 216 billion) 
particles on Ranger at the Texas Advanced Computing Centre (SunBlade 
x6420 with AMD x86\_64 Opteron Quad Core 2300 MHz (9.2 GFlops)
``Barcelona'' processors and Infiniband networking, a total of 62,976 
computing cores and 125,952 GB of total memory. Its shared-memory nodes 
consist of 4 Quad Core processors and 32 GB of RAM.) and JUROPA 
at J\"ulich Supercomputing Centre in Germany (Intel Xeon X5570 
(Nehalem-EP) quad-core, 2.93 GHz, Infiniband networking, a total of 17,664
computing cores and 52 TB of total memory. Its shared-memory nodes 
consist of 2 Quad Core processors and 24 GB of RAM). Since it is not 
practical to perform dedicated scaling tests on such large number 
computing cores, we instead use data extracted from production runs, 
as listed in Table \ref{summary_scaling_table2}. We have found the 
code to scale quite well, within 1.5\% of ideal up to 4,096 cores. 
For very large number of cores (10,976) the scaling is slightly worse,
but still within $\sim20\%$ of ideal, due to increased communication 
costs, I/O overheads (a single timeslice of $5488^3$ particles is 3.6 TB)
and load balancing issues. These first three Ranger runs were performed 
with 4 MPI processes and 4 threads per Ranger node ('4way') (using the 
specially-provided {\it tacc\_affinity} NUMA script to bind the memory 
usage to local core, thus ensuring memory affinity. This is important 
because the 32 GB RAM/node on Ranger (and other computers with 
multi-processor sockets) are actually not equal-access and the local 
8 GB memory of each processor has much shorter access time).

Furthermore, due to the very strong clustering of structures at those small 
scales, some of the cuboid sub-domains came to contain well above the 
average number of particles, thereby requiring more memory per MPI process
in order to run. As a consequence, throughout most of the evolution the 
largest two of these simulations were run with 4096 and 21,952 cores and 
with only 2 MPI processes and 8 threads per Ranger node (2way), which on 
Ranger allows using up to 16 GB of RAM per MPI process. In order to insure 
local memory affinity a special NUMA control script (which we called {\it 
tacc\_affinity\_2way}) 
was developed for us by TACC staff and was used by us to run more efficiently 
in this mode. However, this not-fully-local memory access does affect the 
performance somewhat, as does the imperfect load balancing in such situations,
as seen by the rightmost point on the scaling plot. The scaling in this case
is 42\% below the ideal, although we note that we still get $\sim1.5$ speedup
from doubling the core count, even given these issues. Overall the
code scaling performance is thus quite satisfactory and it clearly 
scales very well to extremely large number of cores and can therefore 
do very large problems efficiently, utilizing the next generation 
Petascale systems.

Finally, we note that several special fixes had to be developed by TACC 
and JUROPA staff in order for our largest runs work properly and to be 
able to use the software libraries we required (MPICH and FFTW), as those 
latter exhibited serious problems when applied to runs of such 
unprecedented size.

### 2012-05-08 Systematics.tex
Updated references, Intro, scaling, halos with Ilian's input.

\section{Systematics}
Random shift, resolution, initial redshift

>>>>

\section{Accuracy}
\label{sec:accuracy}

One of the most accurate way to assess the simulation's accuracy at evolving particles
is to measure the density power spectrum at late time, and compare to non-linear prediction. 
For an over-density field $\delta({\bf \it x})$, the power spectrum is extracted from the two point function in Fourier space as:
\begin{eqnarray}
\langle | \delta ({\bf \it k}) \delta ({\bf \it k'}) | \rangle = (2\pi)^{3}P(k)\delta_{D}({\bf \it k'} - {\bf \it k})
\label{eq:power}
\end{eqnarray}
where the angle bracket corresponds to an ensemble (or volume) average.
We plot in Fig. \ref{fig:power_highres} the dimensionless power spectrum for a large run ($4000^3$ particles),
which are evolved from a Gaussian initial distribution at $z=100$ down to $z=0$, in a box $3.2 h^{-1}\mbox{Gpc}$ per side.
We observe that the agreement with the non-linear prediction of \cite{Lewis} is at the per cent level over a large dynamical range.
The drop of power at high-$k$ is caused by the finite resolution, and the fluctuations at low-$k$ is caused by the white noise imposed
in the initial conditions.

\begin{figure*}%[ht]
  \begin{center}
    \includegraphics[width=3.2in]{graphs/power_highres.eps}
  \caption{Dark matter power spectrum, measured at $z=0$ in a volume $3.2 h^{-1}\mbox{Gpc}$ per side,
  from $4000^3$ particles. The simulation ran on 4000 cores on RANGER.
    \label{fig:power_highres}}
\end{center}
\end{figure*}

(Show plots of pairwise force (transverse and radial), density force, 
mention the work on covariance matrix by Ngan, Harnoisetal.)

\section{Systematics}
\label{sec:systematics}

To discuss :

2) choice of initial redshift, which can lead to large truncation error if too early, but poor inaccurate Zel'dovich if too late 
{\bf JD, here you could discuss the ra\_max tempering...}

3) limits of resolution caused by softening length,

4) Poisson shot noise if too few particles or unrelaxed systems,

5) finite box size effects (with beat coupling?)

The biggest problem with the pp force calculation is that it 
is anisotropic and depends on the location of the fine mesh with respect 
to the particles. An example are two particles on either side of a grid 
cell boundary. These particles will experience the 1-grid cell separation
NGP mesh force, but if the mesh were shifted such that they were
within the same cell, they would experience the (larger) pp force instead. 
This effect is especially pronounced at the early stages of the simulation where
the density is more homogeneous, and leads to mesh artefacts appearing
in the density field. In order to minimize this systematic effect, 
we randomly shift the particle distribution relative to the mesh by a small
amount -- up to 2 fine grid cells in magnitude -- in each
dimension and at each time step.  This adds negligible computational
overhead as it can be applied during the particle position update
and suppresses the mesh behavior over multiple time steps.
It is possible to shift back the particles at the end of each time steps,
which prevents a random drift of the whole population.
This is convenient if one needs to correlate the initial and final positions of the particles.

We ensure that, on average, this solution balances out the mesh feature,
by tuning the NGP force kernel such as to provide  force as unbalanced as possible at grid cell distances.
This adjustment is performed from the pairwise force test described in section \ref{sec:accuracy}.
